Microxanox: an R package for simulating an aquatic \(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.

### 2023-09-08 inst/manuscript/revision_2/Response_rev2.md

Editor-in-Chief

SoftwareX -->

## Response from Authors to Editor and Reviewer comments

### Introduction

Thank you for the opportunity to revise the manuscript again. We apologise that our first revision was found to be insufficient and that it was not as easy to follow as it should have been.

We provide a guide to the materials we resubmit:

First is this document. In it we use this typeface to give our responses and we use the following typeface when text from the manuscript is being quoted: `this typeface shows text being quoted from the manuscript`. In this document we use the following type of box to indicate comments from the reviewer.

> This kind of box contains comments from the reviewer.

The second document is the revised and clean manuscript, and the third is the revised manuscript with changes shown.

Below, we cover each of the latest reviewer comments in turn.

> It was difficult to follow the [Responses and Suggestions] in the new version. Those sections/paragraphs that have been rewritten, added, or modified in the paper should be included in the reply letter and explicitly highlight the parts that correspond precisely to each remark. The highlighted paper and the clean version should be shown at the end of the reply letter.

Apologies that we provided materials that were difficult to follow. With the current revision we provide two versions of the revised manuscript, as requested, one a clean version, and one with changes highlight. *Rainer, please do this.* We also, as requested, include all revised text in this response document and then reference the location in the revised document by line number. *Rainer, please check that this is so, before resubmission. Wait for our final version of the text before inserting the correct line numbers below.*

> 2.1a-b The comment has been addressed.

Thank you.

> 2.2 The comment has not been addressed.

> Here is the original comment 2.2: `finding steady states that correspond to values of one environmental driver`. The software does not find a steady state. Maybe the authors refer to the software running an open loop of the dynamic system under a set of initial conditions and a subset of parameter values (oxygen diffusivity). Then, the numerical evaluation of steady states could be related to environmental conditions in real systems.

**Response**: We agree--the software does not find a steady state of the system. We no longer claim that it does, and we have added `The software does not provide the user with a steady state.` (lines XXX-YYY). Furthermore, in lines XXX-YYY of the newly revised manuscript we explain that the simulations provide a *final state* of the system, and that this is not guaranteed to be a steady state of the system: `When one wishes to be able to make conclusions about how the *steady state* of the system is affected by the environmental driver, it is very important to note that the *final state* (provided by the simulation) is not guaranteed to be a *steady state*.` (lines XXX-YYY). We also then list the actions that users need to take in order to somewhat safely assume that the provided final state is a steady state of the system: `In order to somewhat safely assume that the final state is a steady state, the user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper @Limberger2023, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation.` (lines XXX-YYY)

> 2.3 The comment has not been addressed.

> Here is the original comment 2.3: `Two methods for finding steady states are implemented`. Again, the software does not find a steady state. Maybe the authors refer to two simulations proposed to evaluate an open loop simulation of the dynamic model numerically.

**Response**: We agree--the software does not find a steady state of the system. We no longer claim that it does, and we have added `The software does not provide the user with a steady state.` (lines XXX-YYY).

> 2.4 The comment has been addressed.

Thank you.

> 2.5 The comment has been addressed.

Thank you.

> 2.6a-b The comment has been addressed.

Thank you.

> 2.6c The comment has not been addressed.

> Here is the original comment 2.6c:  `in order to check if results were sensitive to choice of method`. This sentence needs to be clarified. There is no comparison between numerical evaluations in any sense. The numerical simulations or simulation results are not sensitive to choice or method. Indeed, the numerical solution of the dynamical model is sensitive to a set of parameters and initial conditions. Maybe the authors mean the following: to evaluate the performance of the numerical parametric sensitivity analysis of the dynamical model concerning one parametric variation (oxygen diffusivity) under two schemes.

**Response:** We agree that the software does not itself make a comparison. It only provides users information that they could use to make a comparison. This and the surround text have been revised and the points clarified: `Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow users to compare of their results.` (lines XXX-YYY).

> 2.7 The comment has not been addressed.

> Here is the original comment 2.7: It is not clear why there are two `methods` for numerical parametric sensitivity analysis. The authors should include the arguments of this.

**Response:** We now state that `Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow users to compare of their results.` (lines XXX-YYY). Furthermore, we have added the following text: `An potentially important difference between the two methods is in the system state when a new value of oxygen diffusivity is set. In the replication method, the system state when a new value of oxygen diffusivity is set is always the same. Whereas in the temporal method, the system state when a new value of oxygen diffusivity is set is the final state of the system for the previously set value of oxygen diffusivity. Since some modellers prefer one approach and others another, we decided to implement both.` (lines XXX-YYY)

> 2.8 The comment has not been addressed.

> Here is the original comment 2.8: Please consider rewrite the section 3.3 (see previous 2.1 - 2.7 points) in the sense of Parametric sensitivity analysis, Open loop dynamical behavior, Equilibrium points (see Khalil, H. K. (2002). Nonlinear Systems. Prentice Hall, New Jersey.)

**Response:** in our previous response we attempted to argue that the language and terms used in the text were appropriate and understandable by the target audience (quantitative ecologists with some knowledge of dynamical systems modelling). *Rainer, you could ask an experienced quantitative ecologist for testimony on this and then include that here as evidence that that language is suitable.*

> 3 It wasn't easy to follow the [Suggestion] in the new version.

We interpret this as the reviewer asking for clearer accounts of what was changed, and now provide below, with line numbers corresponding to those in the newly revised document, and the corresponding text in green font. *Rainer, please do this.*

@@

Changed to `to rerun the simulation from`

> Line 119 `The general approach used to find the stable state of the system`

Changed to `The general approach to finding the numerical value of a system's stable state (if it exists)`

> 4 The suggestion seems to have been heeded.

Thank you.

> 5 The suggestion seems to have been heeded.

Thank you.

### 2023-09-08 inst/manuscript/revision_2/manuscript_microxanox_rev2.tex
Reorganisation of folder

  \begin{abstract}
  \(Microxanox\) is an R package to simulate a three functional group ecosystem (cyanobacteria, phototrophic sulfur bacteria, and sulfate-reducing bacteria) with four chemical substrates (phosphorus, oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary differential equations. Simulations can be run individually or over a parameter range. The model can be implemented with different numbers of species per functional group. The package is constructed in such a way that the results contain the input parameter used, so that a saved results can be loaded again and the simulation be repeated. Furthermore, the package framework and code should serve as a useful starting point for making simulation models of other types of ecosystems.
  \end{abstract}
    \begin{keyword}
    reproducibility \sep regime shift \sep steady state \sep final state \sep 
    ordinary differential equations
  \end{keyword}

@@

\hypertarget{required-metadata}{%
\section{Required Metadata}\label{required-metadata}}

\hypertarget{current-code-version}{%
\subsection{Current code version}\label{current-code-version}}

@@

\hypertarget{motivation-and-significance}{%
\section{Motivation and significance}\label{motivation-and-significance}}

@@

\hypertarget{software-description}{%
\section{Software description}\label{software-description}}

@@

\hypertarget{software-architecture}{%
\subsection{Software architecture}\label{software-architecture}}

@@

\hypertarget{running-one-simulation}{%
\subsection{Running one simulation}\label{running-one-simulation}}

@@

\hypertarget{examining-ecosystem-responses-to-environmental-change}{%
\subsection{Examining ecosystem responses to environmental change}\label{examining-ecosystem-responses-to-environmental-change}}

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's steady state (if it exists). The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity.

When one wishes to be able to make conclusions about how the steady state of the system is affected by the environmental driver, it is very important to note that the \emph{final state} (found by the simulation) is not guaranteed to be a \emph{steady state}. The user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper \citet{Limberger2023}, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow comparison of their results. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the \emph{Replication method}). This is the method used in the \citet{Bush2017} study and was used to obtain the results in figures 3 and 4 of that article.

@@

\hypertarget{analysing-and-visualising-results}{%
\subsection{Analysing and visualising results}\label{analysing-and-visualising-results}}

@@

\hypertarget{impact}{%
\section{Impact}\label{impact}}

@@

\hypertarget{use-case-1-regime-shifts-during-temporal-environmental-change}{%
\subsection{Use case 1: Regime shifts during temporal environmental change}\label{use-case-1-regime-shifts-during-temporal-environmental-change}}

@@

{\centering \includegraphics[width=350px]{figures/gen_uc1_partrep_temporal_state_switching}

@@

\hypertarget{use-case-2-the-extent-of-hysteresis-depends-on-community-composition}{%
\subsection{Use case 2: The extent of hysteresis depends on community composition}\label{use-case-2-the-extent-of-hysteresis-depends-on-community-composition}}

The package contains a function to extract summary features of ecosystem responses to environmental change, such as the amount of hysteresis displayed by the ecosystem. Hysteresis is a key feature of ecosystem responses to environmental change, because it is related to how difficult it can be to reserve the effects of environmental change \citep{Scheffer2001}. The amount of hysteresis is measured as the extent of the environmental condition (here oxygen diffusitivity) for which there were two steady states. I.e., it is the extent of the environmental conditions for which historical conditions play an important role in determining the current system state.

@@

{\centering \includegraphics[width=150px]{figures/gen_uc2_user_guide_hysteresis}

@@

\hypertarget{use-case-3-effects-of-functional-diversity-on-regime-shifts}{%
\subsection{Use case 3: Effects of functional diversity on regime shifts}\label{use-case-3-effects-of-functional-diversity-on-regime-shifts}}

@@

\hypertarget{adapting-for-other-ecosystems-and-organisms}{%
\subsection{Adapting for other ecosystems and organisms}\label{adapting-for-other-ecosystems-and-organisms}}

@@

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

@@

\hypertarget{conflict-of-interest}{%
\section{Conflict of Interest}\label{conflict-of-interest}}

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

>>>>

  \begin{abstract}
  \(Microxanox\) is an R package to simulate a three functional group ecosystem (cyanobacteria, phototrophic sulfur bacteria, and sulfate-reducing bacteria) with four chemical substrates (phosphorus, oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary differential equations. Simulations can be run individually or over a parameter range. The model can be implemented with different numbers of species per functional group. The package is constructed in such a way that the results contain the input parameter used, so that a saved results can be loaded again and the simulation be repeated. Furthermore, the package framework and code should serve as a useful starting point for making simulation models of other types of ecosystems.
  \end{abstract}
    \begin{keyword}
    reproducibility \sep regime shift \sep final state \sep 
    ordinary differential equations
  \end{keyword}

@@

\section{Required Metadata}\label{required-metadata}

\subsection{Current code version}\label{current-code-version}

@@

\section{Motivation and significance}\label{motivation-and-significance}

@@

\section{Software description}\label{software-description}

@@

\subsection{Software architecture}\label{software-architecture}

@@

\subsection{Running one simulation}\label{running-one-simulation}

@@

\subsection{Examining ecosystem responses to environmental change}\label{examining-ecosystem-responses-to-environmental-change}

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's state. The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one can discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity. In terms of non-linear systems analysis, this would be termed a \emph{parameteric sensitivity analysis} being conducted by running an open loop of the dynamic system under a set of initial conditions and a subset of parameter values (here oxygen diffusivity).

When one wishes to be able to make conclusions about how the \emph{steady state} (or an equilibrium point) of the system is affected by the environmental driver, it is very important to note that the \emph{final state} (provided by the simulation) is not guaranteed to be a \emph{steady state}. The software does not provide the user with a steady state. In order to somewhat safely assume that the final state is a steady state, the user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper \citet{Limberger2023}, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow users to compare of their results. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the \emph{Replication method}). This is the method used in the \citet{Bush2017} study and was used to obtain the results in figures 3 and 4 of that article.

@@

\subsection{Analysing and visualising results}\label{analysing-and-visualising-results}

@@

\section{Impact}\label{impact}

@@

\subsection{Use case 1: Regime shifts during temporal environmental change}\label{use-case-1-regime-shifts-during-temporal-environmental-change}

@@

{\centering \includegraphics[width=350px]{../figures/gen_uc1_partrep_temporal_state_switching}

@@

\subsection{Use case 2: The extent of hysteresis depends on community composition}\label{use-case-2-the-extent-of-hysteresis-depends-on-community-composition}

The package contains a function to extract summary features of ecosystem responses to environmental change, such as the amount of hysteresis displayed by the ecosystem. Hysteresis is a key feature of ecosystem responses to environmental change, because it is related to how difficult it can be to reserve the effects of environmental change \citep{Scheffer2001}. The amount of hysteresis is measured as the extent of the environmental condition (here oxygen diffusitivity) for which there were two distinct final states. I.e., it is the extent of the environmental conditions for which historical conditions play an important role in determining the current system state.

@@

{\centering \includegraphics[width=150px]{../figures/gen_uc2_user_guide_hysteresis}

@@

\subsection{Use case 3: Effects of functional diversity on regime shifts}\label{use-case-3-effects-of-functional-diversity-on-regime-shifts}

@@

\subsection{Adapting for other ecosystems and organisms}\label{adapting-for-other-ecosystems-and-organisms}

@@

\section{Conclusions}\label{conclusions}

@@

\section{Conflict of Interest}\label{conflict-of-interest}

@@

\section{Acknowledgements}\label{acknowledgements}
